---
title: "Homework 02"
author: "Konrat Pekkip"
output: github_document
---

```{r load-packages, message = FALSE, include = FALSE}
#load required packages
library(tidyverse)
library(fs)
library(ggridges)
library(dsbox)
library(janitor)
library(stringr)
library(scales)
library(cowplot)
```

1.  **A new day, a new plot, a new geom.**

```{r exercise-1}
#create ridge density plot of airbnb review scores by neighorhood
edibnb %>%
  group_by(neighbourhood) %>%
  mutate(median_review = median(review_scores_rating, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(neighbourhood = fct_reorder(.f = neighbourhood, .x = median_review)) %>%
  ggplot(data = subset(., !is.na(neighbourhood)),
         mapping = aes(x = review_scores_rating, 
                       y = neighbourhood,
                       fill = neighbourhood))+
  geom_density_ridges(na.rm = TRUE,
                      quantile_lines=TRUE,
                      quantile_fun=function(x,...)median(x),
                      show.legend = FALSE)+
  theme_minimal()+
  labs(title = "Distribution of AirBnB Review Scores by Neighborhood of Edinburgh",
       subtitle = "In Descending Order Based on Median Review Score",
       x = "Density of AirBnB Review Scores",
       y = "Neighborhood",
       caption = "Source: dsbox Package in R")
```

The above plot showcases the distribution and density of AirBnb review scores by neighborhood for the city of Edinburgh in Scotland. The vertical lines indicate the median review score of AirBnBs in a given neighborhood, by which the neighborhoods are also sorted along the y axis. Morningside and Bruntsfield appear to be the neighborhoods with the most popular median AirBnB offerings, as their median ratings are the highest. The distribution of AirBnB reviews appears to differ by neighborhood as well; some neighborhoods (e.g. Tollcross and Southside) have a greater left-skewed tail of the distribution, indicating a greater propensity towards (relatively) lower review scores. Overall, I think the distribution of AirBnb review scores is rather similar across neighborhoods, which might indicate that there is not much of a qualitative difference in AirBnB offerings by neighborhood of Edinburgh. It would be interesting to compare Edinburgh to other cities like Glasgow or Dundee to get a better understanding whether this is a broader trend across Scottish cities, or specific to Edinburgh.

2.  **Key lyme pie.**

```{r exercise-2, warning = FALSE}
#create data frame
lyme <- tribble(
  ~"state", ~"case_count",
  "Pennsylvania", 10208,
  "New Jersey", 4000,
  "New York", 3638,
  "Wisconsin", 1869,
  "Connecticut", 1859,
  "Maine", 1405,
  "Minnesota", 1541,
  "New Hampshire", 1428,
  "Maryland", 1382,
  "Virginia", 1139,
  "Rhode Island", 1111,
  "West Virginia", 671,
  "Vermont", 576,
  "Delaware", 520,
  "Ohio", 293,
  "Remaining States + DC", 2026) %>%
  mutate(case_percent = round((case_count / sum(case_count)) * 100, digits = 1)) %>%
  mutate(state = as.factor(state))

#create pie chart
lyme %>%
  ggplot(mapping = aes(x = 1, y = case_count, fill = fct_relevel(fct_reorder(state, case_count, .desc = FALSE), "Remaining States + DC", after = 0))) +
  geom_bar(stat = "identity") +
  coord_polar(theta = "y", start=0) +
  labs(title = "Lyme Disease Association Lyme Disease Analysis\n2018 US Reported Lyme Disease Cases Featuring Top 15 States",
       fill = NULL,
       caption = "U.S. TOTAL CASES 2018: 33,666\nSource data compiled from CDC pub. data (DVBD)") +
  scale_fill_viridis_d(option = "turbo") +
  guides(fill = guide_legend(reverse=TRUE)) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.title.position = "plot",
        legend.box.background = element_rect(colour = "black"),
        legend.key.size = unit(0.4, "cm"),
        panel.background = element_rect(fill = "white", size = 1),
        plot.caption = element_text(hjust = -0.1))

#present information as bar graph
```

```{r}
#to be deleted:

+
 panel_border(color = "black", size = 1, linetype = 1, remove = FALSE)

,
        panel.border = element_rect(color = "black")

        plot.caption = element_text(hjust = c(0, 0), vjust = -2),


        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.box.background = element_rect(colour = "black")
```

3.  **Foreign Connected PACs.**

```{r exercise-3-clean-data, include = FALSE}
# get a list of files with "Foreign Connected PAC" in their names
list_of_files <- dir_ls(path = "data", regexp = "Foreign Connected PAC")

# read all files and row bind them
# keeping track of the file name in a new column called year
pac <- read_csv(list_of_files, id = "year")

#clean data and pivot_longer the dems/repubs variables
cleaned_pac <- pac %>%
  clean_names() %>%
  separate(col = country_of_origin_parent_company,
           into = c("country_of_origin", "parent_company"),
           sep = '/') %>%
  separate(col = year,
           into = c("remove","year"),
           sep = "-") %>%
  select(-remove) %>%
  mutate(year = as.numeric(str_remove(string = year,
                           pattern = ".csv"))) %>%
  pivot_longer(cols = c("dems", "repubs"),
               names_to = "party",
               values_to = "contribution_amount") %>%
  mutate(contribution_amount = as.numeric(str_remove(string = contribution_amount,
                                                     pattern = "\\$")),
         total = as.numeric(str_remove(string = total,
                                       pattern = "\\$")))
```

```{r exercise-3-visualizations}
#recreate visualization for contributions from UK companies
cleaned_pac %>%
  filter(country_of_origin == "UK") %>%
  group_by(year, party) %>%
  summarize(totalsum = sum(contribution_amount)) %>%
  mutate(party = ifelse(party == "dems", "Democrat", "Republican"))  %>%
  ungroup() %>%
  group_by(party) %>%
  ggplot(mapping = aes(x = year, y = totalsum, color = party), na.rm = TRUE) +
  geom_line(size = 1.5) +
  scale_color_manual(values = c("blue", "red")) +
  scale_x_continuous(name = "Year", 
                     breaks = seq(2000, 2020, 4)) +
  scale_y_continuous(name = "Total Amount",
                     labels = label_number(prefix = "$",
                                           suffix = "M", 
                                           scale = 1e-6)) +
  theme_minimal() +
  theme(axis.title.y = element_text(hjust = 0),
        axis.title.x = element_text(hjust = 0),
        legend.justification=c(1,0), 
        legend.position=c(1,0)) +
  labs(title = "Contributions to US political parties from UK-connected PACs",
       color = "Party",
       caption = "Source: OpenSecrets.org")

#create similar visualization for contributions from another country (Germany)
cleaned_pac %>%
  filter(country_of_origin == "Germany") %>%
  group_by(year, party) %>%
  summarize(totalsum = sum(contribution_amount)) %>%
  mutate(party = ifelse(party == "dems", "Democrat", "Republican"))  %>%
  ungroup() %>%
  group_by(party) %>%
  ggplot(mapping = aes(x = year, y = totalsum, color = party), na.rm = TRUE) +
  geom_line(size = 1.5) +
  scale_color_manual(values = c("blue", "red")) +
  scale_x_continuous(name = "Year", 
                     breaks = seq(2000, 2020, 4)) +
  scale_y_continuous(name = "Total Amount",
                     labels = label_number(prefix = "$",
                                           suffix = "M", 
                                           scale = 1e-6)) +
  theme_minimal() +
  theme(axis.title.y = element_text(hjust = 0),
        axis.title.x = element_text(hjust = 0),
        legend.justification=c(1,0), 
        legend.position=c(1,0)) +
  labs(title = "Contributions to US political parties from Germany-connected PACs",
       color = "Party",
       caption = "Source: OpenSecrets.org")
```

The above visualizations showcase the amount of contributions, in dolllars, that companies from the United Kingdom and Germany have issued to the two major American political parties, the Democrats and the Republicans. Companies from both countries appear to support the Republican party more so than the Democratic party, perhaps due to the Republicans' laissez-faire approach to regulating (big) corporations. We can also see that for companies from both the UK and Germany, the amount of money donated to American parties has increased over time, albeit with ups and downs. Perhaps this is related to court rulings like *Citizens United vs FEC* that make it easier for companies to donate money to political parties in the US.

Whereas to a degree in the UK, and more so in Germany, recent governments have been closer, both ideologically and policy-wise, to the Democratic party than the Republicans, the same does not appear to hold true for companies based in these countries. However, this might have changed during the Trump administration: we can see that both companies in the United Kingdom and companies in Germany increased their contributions to the Democratic party leading up to the critical 2020 Presidential election whereas their contributions to the GOP decreased. Perhaps corporations in both countries were worried about increasing political instability in the United States to the degree where these concerns outweighed the economic benefits that came with supporting the Republican party. It might also be due to the tariffs the Trump administration placed on foreign companies during the trade war with the European Union. Perhaps foreign corporations tend to simply support the party they view most likely not to support a protectionist approach to economic policy.

4.  **Hop on.**

