---
title: "Homework 04"
author: "Konrat Pekkip"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(babynames)
library(ggtext)
```

1.  **How popular am I?**

```{r exercise-1}
#load data
babydata <- read_csv("data/babynames.csv", show_col_types = FALSE)

#visualize popularity of my name
babydata %>%
  filter(name == "Konrad" | name == "Conrad",
         sex == "M") %>%
  ggplot(mapping = aes(x = year, y = n, color = name)) +
  geom_line() +
  scale_x_continuous(breaks = seq(1880, 2020, by = 20)) +
  scale_y_continuous(breaks = seq(0, 600, by = 100)) +
  scale_color_manual(values = c("deepskyblue", "navy")) +
  theme_minimal() +
  theme(plot.title = element_markdown(),
        plot.subtitle = element_textbox_simple()) +
  labs(title = "<span style = 'color:deepskyblue'>Number of Babies Named Konrad or Conrad</span>",
       subtitle ="<span style = 'color:navy'>In the United States between 1880 and 2020</span>",
       x = "Year",
       y = "Number of Babies Named Konrad or Conrad",
       color = "Name")
```

My name, Konrat, is unavailable in the dataset. This not particularly surprising to me, as it is a rather uncommon Turkish name. My parents gave me this name in part because people in Germany (where I grew up) would be familiar with the similar German name "Konrad," and perhaps more accepting of me than if I had a more recognizably Turkish name like Ahmet (my father's name) or Can. Therefore, even though it's not my name, I decided to browse the dataset for the names "Konrad" and "Conrad." It appears as though in the United States, the spelling "Conrad" has always been more popular than "Konrad," and I even know of some famous American "Conrads" -- e.g. Conrad Hilton. Most of the famous "Konrads" I'm aware of are German; take, for example, Konrad Zuse (a pioneer in computer science) or Konrad Adenauer (first post-WWII Chancellor of Germany).


2.  **The app**



3.  **The styled app**

```{r label-me-3}

```
