---
title: "Homework 01"
author: "Konrat Pekkip"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(knitr)
library(dsbox)
library(openintro)
opts_chunk$set(echo=TRUE, include = TRUE, warning=FALSE, message=FALSE)
```

## Question 1: Reverse-Engineering the Grammar of Graphics

**Data**: The data underlying this graph comes from ["The Covid Tracking Project"](https://covidtracking.com/).

**Aesthetics**: The creators of the graph map the following aesthetics to the data: Percent change in the 7-day average of tests administered (mapped to the y-axis, indicated by the color green), and percent change in the 7-day average of new COVID infections recorded (mapped to the y-axis, indicated by the color orange^[I'm not sure if this is accurate as I'm partially colorblind.]), both of which are relative to the starting value from June 1st, 2020. A third aesthetic mapped to the data is time (mapped to the x-axis), namely the time period between June 1st, 2020 and July 5th, 2020.

**Geometries**: The creators of the graph utilize line graphs to visualize the trends in the dataset. Moreover, they highlight the differences between the two variables mapped to the y-axis, number of tests administered and cases recorded, by shading the area between the two lines.

**Facets**: The authors analyze the data not for the whole United States, but faceted by states, allowing for comparisons between every state (except Vermont and Hawai'i). Moreover, the graph is split in two groups, one of which consists of the 28 states with higher numbers of new cases than testing, the other consisting of the 21 states with higher numbers of tests than new cases.

**Statistics**: The authors chose to include in the graph the percentage changes in numbers for both the number of new cases as well as the number of tests administered. This is particularly helpful because they do not include the x and y axes (and corresponding values) in the visualization.

**Coordinates**: The authors chose not to include x or y axes in the visualization of the dataset, and do not provide another coordinate system in the visualization of the data.

**Theme**: The authors chose to visualize the data using a very simplistic theme that does not include a coordinate system (as mentioned before) or a grid that would help us better understand trends in the data.

## Question 2: Road traffic accidents in Edinburgh

```{r exercise-2}

```

## Question 3: NYC marathon winners

```{r exercise-3a}
#define dataset of interest
marathon <- nyc_marathon

#create histogram of the distribution of marathon times
ggplot(data = marathon,
       mapping = aes(y = time_hrs),)+
  geom_histogram(binwidth = 0.03, na.rm = TRUE, fill = "navy")+
  theme_minimal()+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners",
       "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       x = "Number of Winners Finishing per Completion Time",
       caption = "Source: OpenIntro Package")

#create box plot of the distribution of marathon times
ggplot(data = marathon,
       mapping = aes(y = time_hrs),)+
  geom_boxplot(na.rm = TRUE, color = "navy")+
  theme_minimal()+
  theme(axis.text.x=element_blank())+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       caption = "Source: OpenIntro Package")
```

The above histogram and box plots show us the distribution of finishing times of NYC Marathon winners between 1970 and 2020. The box plot indicates that the interquartile range of these finishing times is between a little over 2.125 and a little under 2.5, with the median finishing time being roughly 2.375 hours. We can also see that there are four outliers in the data, all of which Marathon winners that won with relatively slow times. Neither the median nor the interquartile range are apparent in the histogram. However, the histogram potentially indicates the mode of the data, though no conclusive statements can be made about the mode as the visualization of it depends on the bin width which is chosen arbitrarily. The histogram also provides more insights into the distribution of values within the interquartile range; for example, we can tell that the distribution of winning times is bimodal (perhaps due to gender?). The box plot does not provide us with those same insights into the data structure.

```{r exercise-3b}
#create side by side box plots of finishing time by gender
ggplot(data = marathon,
       mapping = aes(y = time_hrs, color = division),)+
  scale_color_manual(values = c("magenta", "cyan"))+
  geom_boxplot(na.rm = TRUE, show.legend = FALSE)+
  facet_wrap(vars(division))+
  theme_minimal()+
  theme(axis.text.x=element_blank())+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners by Gender",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       x = "Gender",
       caption = "Source: OpenIntro Package")
```

Comparing the completion times of female (cyan-colored) and male (magenta-colored) NYC Marathon winners, one can see that men tend to have lower completion times and run the Marathon quicker than women, indicated by the lower median and interquartile range. However, there are a few outliers among the male runners who appear to have finished the race at a similar or slower time to the median of female Marathon winners. In comparison to the box plot constructed in exercise 3a, the box plots divided by gender reveal more outliers. 

```{r exercise-3c}
ggplot(data = marathon,
       mapping = aes(y = time_hrs, color = division))+
  scale_color_manual(values = c("magenta", "cyan"))+
  geom_boxplot(na.rm = TRUE, show.legend = FALSE)+
  theme_minimal()+
  theme(axis.text.x=element_blank())+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners by Gender",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       x = "Gender",
       caption = "Source: OpenIntro Package")
```

[INCLUDE EXPLANATION OF WHAT'S REDUNDANT]

```{r exercie-3d}
ggplot(data = marathon,
       mapping = aes(x = year, y = time_hrs, color = division, shape = division))+
  scale_color_manual(values = c("magenta", "cyan"), name = "Gender")+
  scale_shape_manual(values = c(16, 17), name = "Gender")+
  geom_point(na.rm = TRUE)+
  theme_minimal()+
  labs(title = "Finishing Times of NYC Marathon Winners by Year and Gender",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       x = "Year",
       y = "Marathon Completion Time in Hours",
       caption = "Source: OpenIntro Package")
```

This plot is the first one that allows us to better understand the changes in completion times by gender *and* over the years. We can see that every year, the male winner of the NYC marathon reached the finish line before the female winner of that year. Moreover, we can see that in some years, both the male and female winner finished quicker and in others, both finished slower. This might indicate that performance depends on situational factors that differ by year (e.g. different weather). We can also see that starting in the first few years, i.e. the early 1970s, both men and women performed worse in the Marathon. Starting in the mid-1970s, the completion times appear to stabilize for both men and women. This might be due to technological advances (e.g. new shoe technology) that both male and female participants benefited starting in the mid-1970s.

## Question 4: US counties

```{r exercise-4}

```

## Question 5: Napoleon’s march.

```{r exercise-5}

```

