---
title: "Project 1 Proposal"
author: "Elated Anura"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
```

## Dataset

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.
```{r}
# The dataset we’re analyzing is available on kaggle and  contains historical data on the modern Olympic game and includes all games starting from Athens 1896 till 2016. This data was scraped from www.sports-reference.com in 2018 and wrangled by rgriffin. Till 1992 winter and summer games used to be held in the same year. After 1992, both summer and winter games take place in a 4-years cycle such that winter game in 1994, summer in 1996, then again winter in 1998 and so on.

#This data contains 271116 observations and 15 variables. Each row contains data on an individual athlete participating in an individual olympic athlete event. The name of columns are:

#ID - A unique number for each athlete
Name - Name of athlete
Sex - M or F, M for male, F for Female
Age - An integer showing the age of athlete in years
Height - Height of athlete In centimeters
Weight - Weight of athlete in kilograms
Team - Team name/Name of the country an athlete represents
NOC - National Olympic Committee 3-letter code
Games - Year and season in which the game occurs
Year - An integer showing the year in which the event takes place
Season - Summer or Winter
City - Name of city that hosts the olympic game
Sport - Name of the sport played
Event - Name of particular event
Medal - Gold, Silver, Bronze, or NA

```


```{r}
olympics_data <- read.csv("/Users/kamranahmed/Documents/GitHub/proj1-elated-anura/data/olympics.csv")
```

```{r}
head(olympics_data)
```
```{r}
olympics_data$sex_dummy <- ifelse(olympics_data$sex=="M", 1,0)
vars_of_interest <- c("sex_dummy", "age", "height", "weight")
library(tidyverse) # data manipulation
library(psych) # describe function: statistical summary
library(broom) # tidy function: statistical outputs into tibbles
library(kableExtra) # make-up tables
library(stargazer) # nice regression outputs
olympics_data %>%
  select(vars_of_interest) %>%
  lapply(., function(i) mean(i)) %>% 
  do.call(rbind, .) %>%
  kable(digits = 3)
```

## Questions

The two questions you want to answer.

```{r}


```

## Analysis plan

A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
