---
title: "Olympics and Society -- Cold War Politics and Gender Equality in the Arena"
subtitle: "US vs USSR, Females relative to Males"
author: "Elated Anura <br> Xinyu, Kamran, Konrat"
institute: "University of Chicago"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
# Remove any packages from this list that you're not using
library(tidymodels)
library(knitr)
library(xaringanthemer)
library(tidyverse) # data manipulation
library(psych) # describe function: statistical summary
library(broom) # tidy function: statistical outputs into tibbles
library(kableExtra) # make-up tables
library(stargazer) # nice regression outputs
library(here) # simplifies relative file paths
library(countrycode) #get country names
library(emojifont) # add emoji font
library(cowplot) # combine plot
library(ragg)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, 
  dpi = 300, 
  fig.width = 6, 
  fig.asp = 0.714, 
  out.width = "70%"
  )
```

```{r load-data, include=FALSE}
#read in datasets
olympics_data <- read_csv(here("data", "olympics.csv"))
```

```{r style-slides, echo = FALSE}
style_mono_light(base_color = "#800000")
```

# Introduction
--

## 1. Cold War Politics


--

- How has the dominance of the Soviet Union (/Russia as its main successor) and the United States changed throughout the cold war and after?
--
-


- How do NATO and Warsaw Pact member-states compare in terms of medals won at the Olympics?
--
-







## 2. Gender Equality
--



- How has female participation changed in Olympic Games from 1900 to 2016?
--
-


- Which countries have seen highest contributions from females?

--
-



- Data set source: [www.sports-reference.com](www.sports-reference.com)

---

```{r recode-olympics, include=FALSE}
# In this chunk I'm doing a bunch of analysis that I don't want to present 
# in my slides. But I need the resulting data frame for a plot I want to present.

countrycode <- codelist_panel %>%
  rename(noc = ioc,
         country = country.name.en) %>%
  select(noc, continent, country) %>%
  distinct()

#amend countrycode dataset to  fill crucial missing values
countrycode$noc[70] <- "GDR" #east germany
countrycode$continent[70] <- "Europe"
countrycode$noc[51] <- "TCH" #czechoslovakia
countrycode$continent[51] <- "Europe"
countrycode <- rbind(countrycode, c("URS", "Europe", "Soviet Union"))
countrycode <- rbind(countrycode, c("FRG", "Europe", "West Germany"))

#merge olympics and countrycode datasets
olympics_joined <- left_join(olympics_data, unique(countrycode), by = "noc")

#add/manipulate variables to merged dataset
nato <- c("Belgium", "Canada", "Denmark", "France", "Iceland", "Italy", "Luxembourg", 
          "Netherlands", "Norway", "Portugal", "United Kingdom", "United States",
          "Greece", "Turkey", "West Germany", "Germany", "Spain")
wp <- c("Russia", "Hungary", "Bulgaria", "Poland", "German Democratic Republic", 
        "Romania", "Czechoslovakia", "URS", "Soviet Union")
olympics_joined <- olympics_joined %>%
  mutate(medal_binary = ifelse(!is.na(medal), 1, 0),
         medal_weighted = case_when(medal == "Gold" ~ 3,
                                    medal == "Silver" ~ 2,
                                    medal == "Bronze" ~ 1,
                                    TRUE ~ 0),
         nato_wp = case_when(country %in% nato ~ "NATO",
                             country %in% wp ~ "Warsaw Pact",
                             TRUE ~ "Non-Aligned or not applicable"),
         usa_ussr = case_when(noc == "URS" | noc == "RUS" ~ "USSR/Russia",
                              noc == "USA" ~ "USA",
                              TRUE ~ "Other")) %>%
  group_by(games) %>%
  mutate(medal_sum = sum(medal_binary))

#set theme for graphs
bg_theme <- theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = rel(1)),
        axis.title.x = element_text(margin = margin(t = 10), size = rel(0.8)),
        axis.title.y = element_text(margin = margin(r = 10),size = rel(0.8)),
        strip.text = element_text(face = "bold", size = rel(0.8), hjust = 0.5),
        plot.caption = element_text(face = "italic",size = rel(0.8), 
                                    margin = margin(t = 10)),
        legend.title = element_text(face = "bold",size = rel(0.6)),
        legend.text = element_text(size = rel(0.6)))
```


```{r dataframeq1plot1, include=FALSE}

#create data frame with information relevant to create visualization
q1_df <- olympics_joined %>%
  filter(year >= 1922) %>%
  group_by(year, usa_ussr, season) %>%
  summarize(medal_prop = (sum(medal_binary)/medal_sum), .groups = "drop")
```


```{r pq1-analysis-plot-1, echo = FALSE, warning = FALSE}
# Code hidden with echo = FALSE
# Uses data frame from previous chunk
#create plot mapping USA/USSR/others medals over time
q1_df %>%
  ggplot(mapping = aes(x = year, y = medal_prop, color = usa_ussr, na.rm = TRUE))+
  geom_line() +
  geom_vline(xintercept = 1945,
             linetype = "dashed",
             alpha = 0.75) +
  geom_vline(xintercept = 1991,
             linetype = "dashed",
             alpha = 0.75) +
  annotate("text", label = "End of World War II", x = 1945, y = 0.9, size = rel(2.3), hjust = 0) +
  annotate("text", label = "Dissolution of the USSR", x = 1991, y = 0.5, size = rel(2.3), hjust = 1.01) +
  facet_wrap(vars(season)) +
  scale_color_manual(values = c("grey", "blue", "red")) +
  scale_x_continuous(breaks = seq(1920, 2022, 12),
                     name = "Year") +
  scale_y_continuous(breaks = seq(0, 1, 0.2),
                     name = "Proportion of Medals Won") +
  bg_theme +
  theme(axis.text.x = element_text(hjust = 1, size = rel(0.7)),
        panel.spacing = unit(1.5, "lines")) +
  labs(title = str_wrap("Olympic Performance of USSR/Russia and USA over Time"),
       subtitle = "Proportion of Medals Won by Each Country in Summer and Winter Olympics",
       caption = "Source: data from www.sports-reference.com in 2018",
       color = "Country")
```

---

```{r q1-dataframe-plot-2, include=FALSE}
#create function to help recode necessary variables
recode_if <- function(x, condition, ...) {
  if_else(condition, recode(x, ...), x)
}

#create data frame with information relevant to create visualization
q2_df <- olympics_joined %>%
  filter((nato_wp == "NATO" | nato_wp == "Warsaw Pact") & (year >= 1945 & year <= 1991)) %>%
  mutate(country = recode(country, "Germany" = "United Team of Germany (1956-1964)"),
         country = recode(country, "German Democratic Republic" = "East Germany"),
         nato_wp = as.factor(recode_if(nato_wp, country == "United Team of Germany (1956-1964)", "NATO" = "Both"))) %>%
  group_by(country, nato_wp) %>%
  summarize(medals_won = sum(medal_binary), .groups = "drop")

```

```{r q1-plot-2, echo=FALSE, warning=FALSE}

#create stacked bar chart mapping amount of medals won by country
q2_df %>%
  ggplot(mapping = aes(y = reorder(country, medals_won), x = medals_won, fill = nato_wp)) +
  geom_col() +
  scale_fill_manual(values = c("blue", "red", "magenta")) +
  bg_theme +
  labs(title = str_wrap("Medals Won by NATO and Warsaw Pact Countries"),
       subtitle = "In Olympics during the Cold War, 1945-1991",
       x = "Number of Olympic Medals Won",
       y = "Country",
       fill = "Military Alliance") +
  theme(axis.text.y = element_text(size = rel(0.7)))
```

---
```{r Q2-1-female participation, include=FALSE, message=FALSE, warning=FALSE}
# In this chunk I'm creating a data frame that I don't want to present in my slides. But I need the resulting data frame for a plot I want to present.
# Create a summarized table for participation by gender for each season
participate_sex <- olympics_joined %>% 
  group_by(year, season, sex) %>%
  summarize(count_sex = n_distinct(id), 
            count_game = n_distinct(sport), .groups = "drop")%>% 
  group_by(year, season) %>%
  mutate(visit = 1:n()) %>%
  gather("sex", "count_sex", "count_game", key = variable, value = number) %>%
  unite(combi, variable, visit) %>%
  spread(combi, number) %>%
  select(-c("sex_1", "sex_2")) %>% 
  mutate(count_sex_1 = as.numeric(count_sex_1),
         count_sex_2 = as.numeric(count_sex_2),
         count_game_1 = as.numeric(count_game_1),
         count_game_2 = as.numeric(count_game_2)) %>%
  mutate(f_athletes_p = count_sex_1 / sum(count_sex_1, count_sex_2),
         f_games_p = count_game_1 / sum(count_game_1, count_game_2))



```

```{r female participation, echo = FALSE, warning = FALSE, message=FALSE}
# Code hidden with echo = FALSE
# Uses data set from previous chunk
# Plot the time series plot for female participation
q2_p1 <- participate_sex %>% 
  ggplot(aes(x = year, y = f_athletes_p)) +
  geom_line(aes(color = "Number of female athletes")) +
  geom_line(aes(x = year, y = f_games_p, 
                color = "Disciplines females compete in")) + 
  facet_wrap(vars(season), scales = "free") +
  labs(title = str_wrap("Female Athletes Participation in the Olympic Games by Season"),
       subtitle = "From 1900 to 2016",
       x = "Year",
       y = "Proportion",
       caption = "Source: data from www.sports-reference.com in 2018") +
  scale_color_manual(name = "Proportion of",
                     breaks = c("Number of female athletes",  "Games females compete in"),
                     values=c("Number of female athletes" = "orange",
                              "Disciplines females compete in" = "steelblue")) +
  scale_color_manual(name = "Proportion of",
                    breaks = c("Disciplines females compete in", "Number of female athletes"),
                    values=c("Number of female athletes" = "#F4C300",
                             "Disciplines females compete in" = "#0085C7")) +
  bg_theme
q2_p1 
```


---

```{r Q2-2-countryranking_dataframe, dev='ragg_png', include=FALSE, warning=FALSE}
# In this chunk I'm creating a data frame that I don't want to present in my slides. But I need the resulting data frame for a plot I want to present.
# create the sum table for a wide table of each country and its' medals.
medalist_rank <- olympics_joined %>%
  filter(year >= 2000) %>% # focusing on  recent years
  mutate(gold = if_else(medal == "Gold", 1, 0),
         silver = if_else(medal == "Silver", 1, 0),
         bronze = if_else(medal == "Bronze", 1, 0)) %>%
  group_by(country, sex) %>%
  summarize(n_gold = sum(gold, na.rm = TRUE), 
            n_silver = sum(silver, na.rm = TRUE), 
            n_bronze = sum(bronze, na.rm = TRUE), .groups = "drop") %>%
  mutate(n_medal = n_gold + n_silver + n_bronze) %>%
  group_by(country) %>%
  mutate(visit = 1:n()) %>%
  gather("sex", "n_gold", "n_silver", "n_bronze", "n_medal", 
         key = variable, value = number) %>%
  unite(combi, variable, visit) %>%
  spread(combi, number) %>%
  select(-c("sex_1", "sex_2"))
medalist_rank[2:9] = lapply(medalist_rank[2:9], FUN = function(y){as.numeric(y)})
medalist_rank <- medalist_rank %>%
  mutate(fp_medal = n_medal_1 / sum(n_medal_1, n_medal_2))
# rank and use only the top 15 countries for plotting
top_15 <- medalist_rank %>%
  arrange(desc(n_medal_1)) %>%
  head(15) %>%
  mutate(n_b_s = n_bronze_1 + n_silver_1)

```


```{r plot-top15countries, echo = FALSE, warning = FALSE}
# Code hidden with echo = FALSE
# Uses data frame from previous chunk
# Creates plot for top 15 countries with respect to medals won by females
# create two plots with normalized and non-normalized two versions.
q2_p2_1 <- top_15 %>%
  ggplot(aes(x = n_medal_1, y = reorder(country, n_medal_1))) +
  geom_segment(aes(yend = country), xend = 0, colour = "#F4C300", size = 0.5, alpha = 0.6) +
  geom_emoji(alias = 'trophy', x = top_15$n_medal_1, y = top_15$country, 
             size = 3.5, color = "#D6AF36") +
  geom_emoji(alias = 'trophy', x = top_15$n_bronze_1, y = top_15$country, 
             size = 3.5, color = "#824A02") +
  geom_emoji(alias = 'trophy', x = top_15$n_b_s, y = top_15$country, 
             size = 3.5, color = "#A7A7AD") +
  scale_x_continuous(breaks = seq(0, 1000, 200)) +
  annotate("text", label = "Types of medals", x = 600, y = "United Kingdom", size = rel(2.6)) +
  annotate("rect", xmin = 400, xmax = 800, ymin = "Italy", ymax = "Japan", alpha = .1) +
  annotate("point", x = 500, y = "South Korea", colour = "#D6AF36", size = 1) +
  annotate("point", x = 500, y = "Norway", colour = "#A7A7AD", size = 1) +
  annotate("point", x = 500, y = "France", colour = "#824A02", size = 1) +
  annotate("text", x = 600, y = "South Korea", label = "Gold", size = rel(2.4), colour = "#D6AF36") +
  annotate("text", x = 600, y = "Norway", label = "Silver", size = rel(2.4), colour = "#A7A7AD") +
  annotate("text", x = 600, y = "France", label = "Bronze", size = rel(2.4), colour = "#824A02") + 
  labs(x = "Number of Medals Won by Women", y = "Countries",
       caption = " ") +
  theme_minimal() +
  bg_theme +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = rel(0.7))) 
q2_p2_2 <- top_15 %>% 
  ggplot(aes(x = fp_medal, y = reorder(country, n_medal_1))) +
  geom_segment(aes(yend = country), xend = 0, colour = "#009F3D", size = 0.5, alpha = 0.6) +
  geom_emoji(alias = 'trophy', 
             x = c(top_15$fp_medal[15], top_15$fp_medal[3], top_15$fp_medal[7]), 
             y = c("Romania", "China", "Netherlands"),
             size = 4, color = "#009F3D") +
  labs(x = "Proportion of Medals Won by Women", 
       caption = "Source: data from www.sports-reference.com in 2018") + 
  bg_theme +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank()) 
# Adding two plots together for comparison
q2_p2 <- plot_grid(q2_p2_1, NULL, q2_p2_2, rel_widths = c(3.2, -0.7, 3), nrow = 1,
                   align = "hv", labels = c(str_pad("Number", 22, "left", " "), "", 
                                            str_pad("Percent", 23, "left", " ")), label_size = 8)
title <- ggdraw() + 
  draw_label("Countries with the Highest Number of Female Olympic Medalists, 2000-2016",
             fontface = 'bold', x = 0, hjust = -0.1, size = 10) +
  theme(plot.margin = margin(b = 10))
plot_grid(title, q2_p2, ncol = 1, rel_heights = c(0.1, 1))
```



