---
title: "work_q2"
output: html_document
date: '2022-05-29'
---

```{r setup, include=FALSE}
# load packages
library(tidyverse)
library(colorspace)
library(scales)
library(lubridate)
library(data.table)
library(stringi)

# set default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, fig.retina = 3,
  dpi = 400, out.width = "90%"
)
# dplyr print min and max
options(dplyr.print_max = 6, dplyr.print_min = 6)
```
## Import data:
We need only 5 columns from the dataset. Some data exploration shows that the break points columns are `NA` for the datasets from 1990 and before. So, those .csv files can be ignored for our purposes.

```{r}
matches = read_csv("data/atp/atp_matches_2000.csv", col_select = c(-winner_seed,-loser_seed))

## import only the columns you need
# matches = read_csv("data/atp/atp_matches_1991.csv", 
#                    col_types = cols_only(best_of = 'i', w_bpFaced = 'i', w_bpSaved = 'i', l_bpFaced = 'i', l_bpSaved = 'i'))

matches %>% glimpse
```
```{r}
#calculate total successful break points in match
matches = matches %>% mutate(
  total_breaks = (w_bpFaced-w_bpSaved) + (l_bpFaced-l_bpSaved)
)

matches %>% select(best_of, total_breaks) %>% 
  mutate(breaks_per_set = total_breaks/best_of) %>% 
  filter(best_of == 5) %>% 
  summarise(avg_breaks = mean(breaks_per_set, na.rm = TRUE))

```
# Break point data is available for ATP from 1991 onwards, and for WTA from 2003 onwards.  
After dropping `NA`'s, we are left with ~1 million observations (ATP) and ~35k observations (WTA).

Help taken from:
# see: https://stackoverflow.com/questions/5788117/only-read-selected-columns,
# and: https://stackoverflow.com/questions/41355912/extracting-a-numeric-value-from-multiple-filename-in-r

```{r atp_data}
# ----------- read in all ATP match csv files ----------- 
filenames = list.files(path = "./data/atp/", pattern = "atp_matches*", full.names = TRUE) 

combo_data_ATP = map_df(filenames,
                    ~read_csv(.x, 
                              col_types = cols_only(best_of = 'i', w_bpFaced = 'i', w_bpSaved = 'i', l_bpFaced = 'i', l_bpSaved = 'i', surface = "c"), 
                              show_col_types = FALSE) 
                    %>% mutate(year = as.integer(stri_extract_last(.x, regex = "(\\d+)")))
) %>%
  na.omit()

# ----------- calculate stats -----------
breaks_stats_ATP = combo_data_ATP %>% mutate(
  total_breaks = (w_bpFaced-w_bpSaved) + (l_bpFaced-l_bpSaved)) %>%
  mutate(breaks_per_set = total_breaks/best_of) %>% 
  group_by(year, surface) %>% 
  summarise(
    avg_breaks_surface = mean(breaks_per_set, na.rm = TRUE)
    )

breaks_per_year_ATP = breaks_stats_ATP %>% summarise(avg_breaks = mean(avg_breaks_surface))
```
```{r wta_data}
# ----------- read in all WTA match csv files ----------- 
filenames = list.files(path = "./data/wta/", pattern = "wta_matches*", full.names = TRUE) 

combo_data_WTA = map_df(filenames,
                        ~read_csv(.x, 
                                  col_types = cols_only(best_of = 'i', w_bpFaced = 'i', w_bpSaved = 'i', l_bpFaced = 'i', l_bpSaved = 'i', surface = "c"), 
                                  show_col_types = FALSE) 
                        %>% mutate(year = as.integer(stri_extract_last(.x, regex = "(\\d+)")))
) %>%
  na.omit()


# ----------- calculate stats -----------
breaks_stats_WTA = combo_data_WTA %>% mutate(
  total_breaks = (w_bpFaced-w_bpSaved) + (l_bpFaced-l_bpSaved)) %>%
  mutate(breaks_per_set = total_breaks/best_of) %>% 
  group_by(year, surface) %>% 
  summarise(
    avg_breaks_surface = mean(breaks_per_set, na.rm = TRUE)
  )

breaks_per_year_WTA = breaks_stats_WTA %>% 
  summarise(avg_breaks = mean(avg_breaks_surface))

breaks_per_year_WTA
```

```{r combine_and_plot}

# Plot 1
# -------  Combine breaks_per_year datasets for men and women, and plot ----
bind_rows(breaks_per_year_ATP, breaks_per_year_WTA, .id = "id") %>% 
  mutate(id=recode(id, '1'='ATP', '2'='WTA')) %>% 
  ggplot(aes(x=year, y=avg_breaks, colour=id)) +
  geom_point() +  
  geom_hline(yintercept=mean(breaks_per_year_ATP$avg_breaks), colour="red") +
  geom_hline(yintercept=mean(breaks_per_year_WTA$avg_breaks), colour="blue") +
  labs(
    title = "How common are service breaks? (1991-current)",
    y = "Breaks per set",
    x = "Time"
  )

# Plot 2
# ------- Combine combo_data_WTA datasets for men and women, and plot ------
# NB: This has additional `playing surface` information

bind_rows(breaks_stats_ATP, breaks_stats_WTA, .id = "id") %>% 
  mutate(id=recode(id, '1'='ATP', '2'='WTA')) %>% 
  ggplot(aes(x=year, y=avg_breaks_surface, colour=id)) +
  geom_point() +
  facet_wrap(vars(surface)) +
  labs(
    title = "How common are service breaks? (1991-current)",
    subtitle = "Dependence on playing surface",
    y = "Breaks per set",
    x = "Time"
  )


breaks_stats_WTA
```


