---
title: "Homework 01"
author: "Konrat Pekkip"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(knitr)
library(dsbox)
library(openintro)
library(scales)
library(grid)
library(gridExtra)
library(ggrepel)
opts_chunk$set(echo=TRUE, include = TRUE, warning=FALSE, message=FALSE)
```

## Question 1: Reverse-Engineering the Grammar of Graphics

**Data**: The data underlying this graph comes from ["The Covid Tracking Project"](https://covidtracking.com/).

**Aesthetics**: The creators of the graph map the following aesthetics to the data: Percent change in the 7-day average of tests administered (mapped to the y-axis, indicated by the color green), and percent change in the 7-day average of new COVID infections recorded (mapped to the y-axis, indicated by the color orange^[I'm not sure if this is accurate as I'm partially colorblind.]), both of which are relative to the starting value from June 1st, 2020. A third aesthetic mapped to the data is time (mapped to the x-axis), namely the time period between June 1st, 2020 and July 5th, 2020.

**Geometries**: The creators of the graph utilize line graphs to visualize the trends in the dataset. Moreover, they highlight the differences between the two variables mapped to the y-axis, number of tests administered and cases recorded, by shading the area between the two lines.

**Facets**: The authors analyze the data not for the whole United States, but faceted by states, allowing for comparisons between every state (except Vermont and Hawai'i). Moreover, the graph is split in two groups, one of which consists of the 28 states with higher numbers of new cases than testing, the other consisting of the 21 states with higher numbers of tests than new cases.

**Statistics**: The authors chose to include in the graph the percentage changes in numbers for both the number of new cases as well as the number of tests administered. This is particularly helpful because they do not include the x and y axes (and corresponding values) in the visualization.

**Coordinates**: The authors chose not to include x or y axes in the visualization of the dataset, and do not provide another coordinate system in the visualization of the data.

**Theme**: The authors chose to visualize the data using a very simplistic theme that does not include a coordinate system (as mentioned before) or a grid that would help us better understand trends in the data.

## Question 2: Road Traffic Accidents in Edinburgh

```{r exercise-2}
#recreate plot indicating number of accidents during the day
accidents %>%
  mutate(weekend = ifelse(day_of_week == "Saturday" | day_of_week == "Sunday", "Weekend", "Weekday")) %>%
  ggplot(mapping = aes(x = time, fill = severity))+
  geom_density(alpha = 0.5)+
  scale_fill_viridis_d(name = "Severity")+
  facet_wrap(vars(weekend), ncol = 1)+
  theme_minimal()+
  labs(title = "Number of accidents throughout the day",
       subtitle = "By day of week and severity",
       y = "Density",
       x = "Time of Day")
```

The above plot displays the density of accidents that occur in Edinburgh in 2018 both on weekdays and on weekends. The timing and density of slight accidents appears similar on weekends and weekdays, peaking around or a little after 4pm. Similarly, serious accidents appear to peak in the afternoon both on weekends and weekdays. Fatal accidents are only reported to occur on weekdays and not weekends, which is surprising given people's alcohol consumptions on weekends which one could expect to lead to higher numbers of fatal accidents. On weekdays, fatal accidents appear to peak a little before noon, and the number of fatal incidents that occur at night appears higher than those occurring in the afternoon.

## Question 3: NYC Marathon Winners

```{r exercise-3a}
#create histogram of the distribution of marathon times
ggplot(data = nyc_marathon,
       mapping = aes(y = time_hrs))+
  geom_histogram(binwidth = 0.03, na.rm = TRUE, fill = "navy")+
  theme_minimal()+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners",
       "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       x = "Number of Winners Finishing per Completion Time",
       caption = "Source: OpenIntro Package")

#create box plot of the distribution of marathon times
ggplot(data = nyc_marathon,
       mapping = aes(y = time_hrs))+
  geom_boxplot(na.rm = TRUE, color = "navy")+
  theme_minimal()+
  theme(axis.text.x=element_blank())+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       caption = "Source: OpenIntro Package")
```

The above histogram and box plots show us the distribution of finishing times of NYC Marathon winners between 1970 and 2020. The box plot indicates that the interquartile range of these finishing times is between a little over 2.125 and a little under 2.5, with the median finishing time being roughly 2.375 hours. We can also see that there are four outliers in the data, all of which Marathon winners that won with relatively slow times. Neither the median nor the interquartile range are apparent in the histogram. However, the histogram potentially indicates the mode of the data, though no conclusive statements can be made about the mode as the visualization of it depends on the bin width which is chosen arbitrarily. The histogram also provides more insights into the distribution of values within the interquartile range; for example, we can tell that the distribution of winning times is bimodal (perhaps due to gender?). The box plot does not provide us with those same insights into the data structure.

```{r exercise-3b}
#create side by side box plots of finishing time by gender
ggplot(data = nyc_marathon,
       mapping = aes(y = time_hrs, color = division))+
  scale_color_manual(values = c("magenta", "cyan"))+
  geom_boxplot(na.rm = TRUE, show.legend = FALSE)+
  facet_wrap(vars(division))+
  theme_minimal()+
  theme(axis.text.x=element_blank())+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners by Gender",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       x = "Gender",
       caption = "Source: OpenIntro Package")
```

Comparing the completion times of female (cyan-colored) and male (magenta-colored) NYC Marathon winners, one can see that men tend to have lower completion times and run the Marathon quicker than women, indicated by the lower median and interquartile range. However, there are a few outliers among the male runners who appear to have finished the race at a similar or slower time to the median of female Marathon winners. In comparison to the box plot constructed in exercise 3a, the box plots divided by gender reveal more outliers. 

```{r exercise-3c}
#redo previous plot without faceting
ggplot(data = nyc_marathon,
       mapping = aes(y = time_hrs, color = division))+
  scale_color_manual(values = c("magenta", "cyan"))+
  geom_boxplot(na.rm = TRUE)+
  theme_minimal()+
  theme(axis.text.x=element_blank())+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners by Gender",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       x = "Gender",
       color = "Gender",
       caption = "Source: OpenIntro Package")

#redo previous plot without color differences
ggplot(data = nyc_marathon,
       mapping = aes(y = time_hrs))+
  geom_boxplot(na.rm = TRUE, show.legend = FALSE)+
  facet_wrap(vars(division))+
  theme_minimal()+
  theme(axis.text.x=element_blank())+
  labs(title = "Distribution of Finishing Times of NYC Marathon Winners by Gender",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       y = "Marathon Completion Time in Hours",
       x = "Gender",
       caption = "Source: OpenIntro Package")
```

I don't think very many things in the plot from exercise 3b were redundant. The only thing that I noticed is redundant is that I assigning a different color for the different genders *and* made use of faceting to work out the same difference. In the first plot above, I thus redid the same graph but this time without faceting, as the color difference already indicates the gender division. Alternatively, one could facet the plot without using color, which is what I do for the second plot. Neither graph meaningfully changes the data-to-ink ration in my opinion. 

```{r exercie-3d}
ggplot(data = nyc_marathon,
       mapping = aes(x = year, y = time_hrs, color = division, shape = division))+
  scale_color_manual(values = c("magenta", "cyan"), name = "Gender")+
  scale_shape_manual(values = c(16, 17), name = "Gender")+
  geom_point(na.rm = TRUE)+
  theme_minimal()+
  labs(title = "Finishing Times of NYC Marathon Winners by Year and Gender",
       subtitle = "Includes data from NYC Marathons between 1970-2020",
       x = "Year",
       y = "Marathon Completion Time in Hours",
       caption = "Source: OpenIntro Package")
```

This plot is the first one that allows us to better understand the changes in completion times by gender *and* over the years. We can see that every year, the male winner of the NYC marathon reached the finish line before the female winner of that year. Moreover, we can see that in some years, both the male and female winner finished quicker and in others, both finished slower. This might indicate that performance depends on situational factors that differ by year (e.g. different weather). We can also see that starting in the first few years, i.e. the early 1970s, both men and women performed worse in the Marathon. Starting in the mid-1970s, the completion times appear to stabilize for both men and women. This might be due to technological advances (e.g. new shoe technology) that both male and female participants benefited starting in the mid-1970s.

## Question 4: US Counties

```{r exercise-4a}
#examine code from exercise 4a
ggplot(county) +
  #geom_point(aes(x = median_edu, y = median_hh_income)) +
  geom_boxplot(aes(x = smoking_ban, y = pop2017))
```

The code above works in the sense that it produces an output. However, assuming the intention was to create a plot showcasing the relationship between median education level and median household income on one hand, and a box plot indicating the distribution of the population in 2017 by counties with comprehensive, partial or no smoking bans, I believe the above code does not achieve that. 

First of all, I think the coder should have created two separate ggplots, as there is no overlap between variables between the two plots they attempted to create. Secondly, to analyze the relationship between a categorical and a continuous variable (i.e. median education level and median household income), a scatter plot is not the way to go. Instead, the coder could have utilized a box plot to indicate the distribution of household income per education level, or a violin plot in order to indicate the density of observations by income level.

Considering the second plot the coder attempted to create, formally, a box plot is an appropriate way of going about it, as it maps the distribution of a continuous variable (i.e. population in 2017) onto a categorical variable (i.e. smoking ban). However, in this context, I'm not sure what the contribution is, as the two variables do not seem related to me.

```{r exercise-4b}
#recreate plot A from exercise 4b
ggplot(data = county,
       mapping = aes(x = homeownership, y = poverty))+
  geom_point()+
  labs(title = "Plot A")

#recreate plot B from exercise 4b
ggplot(data = county,
       mapping = aes(x = homeownership, y = poverty))+
  geom_point()+
  geom_smooth(se = FALSE,
              method = 'gam',
              formula = y ~ s(x, bs = "cs"))+
  labs(title = "Plot B")

#recreate plot C from exercise 4b
ggplot(data = county,
       mapping = aes(x = homeownership, y = poverty))+
  geom_point(color = "black")+
  geom_smooth(mapping = aes(group = metro),
              se = FALSE,
              color = "green",
              method = 'gam',
              formula = y ~ s(x, bs = "cs"))+
  labs(title = "Plot C")

#recreate plot D from exercise 4b
ggplot(data = county,
       mapping = aes(x = homeownership, y = poverty))+
  geom_point(color = "black")+
  geom_smooth(mapping = aes(group = metro),
              se = FALSE,
              color = "blue",
              method = 'gam',
              formula = y ~ s(x, bs = "cs"))+
  labs(title = "Plot D")

#recreate plot E from exercise 4b
ggplot(data = county,
       mapping = aes(x = homeownership, y = poverty, color = metro))+
  geom_point()+
  geom_smooth(mapping = aes(linetype = metro),
              se = FALSE,
              color = "blue",
              method = 'gam',
              formula = y ~ s(x, bs = "cs"))+
  labs(title = "Plot E")

#recreate plot F from exercise 4b
ggplot(data = county,
       mapping = aes(x = homeownership, y = poverty, color = metro))+
  geom_point()+
  geom_smooth(se = FALSE,
              method = 'gam',
              formula = y ~ s(x, bs = "cs"))+
  labs(title = "Plot F")

#recreate plot G from exercise 4b
ggplot(data = county,
       mapping = aes(x = homeownership, y = poverty, color = metro))+
  geom_point()+
  geom_smooth(mapping = aes(group = 1),
              se = FALSE,
              method = 'gam',
              formula = y ~ s(x, bs = "cs"))+
  labs(title = "Plot G")

#recreate plot H from exercise 4b
ggplot(data = county,
       mapping = aes(x = homeownership, y = poverty, color = metro))+
  geom_point()+
  labs(title = "Plot H")
```

## Question 5: Napoleon’s March.

```{r exercise-5, fig.height=4.7, fig.width=10}
#define data set
napoleon <- read_rds("data/napoleon.rds") #defining an object that contains the list of three tibbles containing all relevant data

#define different datasets from list
cities <- napoleon$cities #defining the cities dataset
temperatures <- napoleon$temperatures #defining the temperatures dataset
troops <- napoleon$troops #defining the troops dataset
breaks <- c(1, 2, 3) * 10^5 #to print values not in exponential but regular notation

#create ggplot with information from troops and cities datasets
ggtroops <- ggplot(data = troops, 
                   mapping = aes(x = long,
                                 y = lat))+ #create ggplot with longitude and latitude as x and y values
  geom_path(aes(size = survivors,
                colour = direction,
                group = group,
                alpha = 0.75), #specifying geom_path aesthetics to include number of survivors, direction (advancing or retreating), and different groups, also change transparency level
            lineend = "round", #round line ends for better visual appeal
            show.legend = FALSE)+ #disable legend
  scale_size("Survivors", range = c(1,10),
             labels = comma(breaks)) + #formatting the scales of the graph
  scale_color_manual("Direction", 
                     values = c("#E8CBAB", "#1F1A1B"),
                     labels=c("Advance", "Retreat"))+ #setting the colors to the ones Minard used in his graph; indicating whether troops were advancing or retreating
  geom_point(data = cities)+ #map city locations onto longitude/latitude
  geom_text_repel(data = cities, 
                  mapping = aes(label = city))+ #to avoid overlaying of different city names and data points
  coord_cartesian(xlim = c(24, 38))+ #set x axis limits to be the same as temperature plot later
  labs(title = "Figurative Map of the Successive Losses in Men of the French Army in the Russian campaign of 1812-1813",
       x = NULL, 
       y = NULL, 
       alpha = NULL)+ #remove axis labels, set title
  guides(color = "none",
         size = "none")+ #removes legends of both the color argument (advance or retreat) and the size argument (number of survivors)
  theme_void()+ #plots graph onto a blank slate as opposed to default grey ggplot2 theme
  theme(plot.title = element_text(hjust = 0.5))

#create ggplot with information from temperature dataset
ggtemp <- temperatures %>%
  mutate(label = paste0(temp, "° ", date))%>% #add ° symbol to temperatures for the graph to be closer to Minard's original
  ggplot(mapping = aes(x = long, y = temp))+ #create ggplot with same longitudinal data as x axis, temperature as y axis
  geom_path(color = "grey", size = 1.5)+ #maps progression of temperature in the color grey
  geom_point(size = 1)+ #specifies size of data points
  geom_text_repel(mapping = aes(label = label), 
                  size=2.5)+ #to make sure all temperature data points are actually displayed on the graph
  coord_cartesian(xlim = c(24, 38))+ #match x axis limits with ggtroops plot
  labs(x = NULL, y="Temperature")+ #set y axis label to be temperature
  theme_bw() + #set theme
  theme(panel.grid.major.x = element_blank(), #removes all grid lines except for the temperature section
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks = element_blank(),#removes axis text and ticks
        panel.border = element_blank(), #removes border around the plotting area
        plot.title = element_text(hjust = 0.5))+ #centers title
  labs(title = "Temperature in Degrees Réaumur",
       caption = "Paris, November 20th, 1869")

#merge both plots into single graph
grid.arrange(ggtroops, ggtemp, nrow=2, heights=c(3.5, 1.2)) #assembles ggtroops and ggtemp plots
grid.rect(width = .99, height = .99, gp = gpar(lwd = 2, col = "gray", fill = NA)) #add border surrounding entire graph
```

In order to recreate Minard's graph covering Napoleon's failed invasion of Russia, I greatly benefited and pulled most of the code from Michael Friendly's blog post titled ["Minard meets ggplot2"](http://euclid.psych.yorku.ca/www/psy6135/tutorials/Minard.html). This was the only external source I used, and it was quite helpful as Friendly provides a step-by-step tutorial (which I followed almost in its entirety!) on how to recreate Minard's famous graph using ggplot2. 

For a detailed explanation of what each line of code above does, please refer to my in-line comments. Broadly, the above chunk of code defines two ggplot elements. The first, `ggtroops`, maps the advances and retreats of Napoleon's army as well as the number of surviving soldiers and geographic location. The lighter, beige-ish color indicates troop movements towards Moscow, and black indicates troops retreating back to France. The second ggplot, `ggtemp`, maps the temperature changes as troops moved closer towards Moscow. I then merged both plots using the `grid.arrange()` function in order to display one graph as the final output.

Some of the changes I made to this plot include adding a title, both for the whole plot as well as for the subsection regarding the temperature changes throughout the campaign. I also slightly decreased the transparency of geom_path in the ggtroops plot, in order to make the plot less overwhelming and to showcase individual datapoints for different cities.
